---
title: "Dental Insurance IBNP Reserving — Case Study"
subtitle: "Completion Factor, Bornhuetter-Ferguson, and Projected Paid Lag Methods"
format: 
  pdf:
    geometry:
      - left=0.75in
      - right=0.75in
      - top=1in
      - bottom=1in
    code-overflow: wrap
    df-print: kable
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This analysis estimates incurred-but-not-paid (IBNP) claim reserves for a
dental insurer using three complementary actuarial reserving methods:

1.  **Completion Factor method** — projects ultimate claims using
    historical payment development patterns
2.  **Bornhuetter-Ferguson (BF) method** — blends a prior expected loss
    ratio with actual paid claims, useful when experience is still
    developing or premium is growing quickly
3.  **Projected Paid Lag method** — estimates ultimate claims from the
    incremental (rather than cumulative) percentage of claims paid at
    each lag

------------------------------------------------------------------------

## Setup

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```

```{r}
# Data is not committed to this repo (proprietary competition dataset).
# Place the competition workbook in the local data/ folder before running:
#   data/case_competition_data.xlsx
data_path <- "../data/case_competition_data.xlsx"

util <- read_excel(data_path, sheet = "Utilization Data")
prem <- read_excel(data_path, sheet = "Premium Data")

colnames(util)[1:3] <- c("service_month", "payment_month", "paid")
colnames(prem)[1:2] <- c("month", "premium")

util <- util |>
  select(service_month, payment_month, paid) |>
  filter(!is.na(service_month), !is.na(payment_month), !is.na(paid)) |>
  mutate(
    service_month = as.Date(service_month),
    payment_month = as.Date(payment_month),
    # lag = number of months between when a claim was incurred and when it was paid
    lag = (as.integer(format(payment_month, "%Y")) - as.integer(format(service_month, "%Y"))) * 12 +
          (as.integer(format(payment_month, "%m"))  - as.integer(format(service_month, "%m")))
  ) |>
  filter(lag >= 0, lag <= 20)

prem <- prem |>
  select(month, premium) |>
  filter(!is.na(month)) |>
  mutate(month = as.Date(month))

eval_date <- as.Date("2023-05-01")
```

------------------------------------------------------------------------

## Build the Cumulative Paid Triangle

```{r}
inc_tri <- util |>
  group_by(service_month, lag) |>
  summarise(paid = sum(paid), .groups = "drop") |>
  pivot_wider(names_from = lag, values_from = paid, values_fill = 0) |>
  arrange(service_month)

lag_cols <- as.character(0:15)
for (col in lag_cols) { if (!col %in% colnames(inc_tri)) inc_tri[[col]] <- 0 }
inc_tri <- inc_tri |> select(service_month, all_of(lag_cols))

# Convert incremental payments to cumulative payments across each lag
cum_tri <- inc_tri
for (i in 2:length(lag_cols)) {
  cum_tri[[lag_cols[i]]] <- cum_tri[[lag_cols[i - 1]]] + cum_tri[[lag_cols[i]]]
}

# Age = how many months old each service month is as of the evaluation date
cum_tri <- cum_tri |>
  mutate(age = (as.integer(format(eval_date, "%Y")) - as.integer(format(service_month, "%Y"))) * 12 +
               (as.integer(format(eval_date, "%m"))  - as.integer(format(service_month, "%m"))))
```

```{r}
# Sanity check: what share of claims are paid by each lag?
util |>
  count(lag) |>
  filter(lag >= 0, lag <= 20) |>
  arrange(lag) |>
  mutate(cumulative_pct = percent(cumsum(n) / sum(n), accuracy = 0.1)) |>
  print(n = 25)
```

------------------------------------------------------------------------

## Method 1: Completion Factor

```{r}
# Treat months 15+ periods old as "mature" (i.e., effectively fully paid)
# and use them to estimate how much of ultimate claims are typically paid
# by each earlier lag.
mature <- cum_tri |> filter(age >= 15)

completion_factors <- sapply(lag_cols, function(l) {
  valid <- mature |> filter(.data[["15"]] > 0, .data[[l]] > 0)
  if (nrow(valid) < 3) return(NA_real_)
  mean(valid[[l]] / valid[["15"]])
})

cf_df <- data.frame(lag = as.integer(lag_cols), cf = completion_factors) |>
  filter(!is.na(cf))
```

```{r}
total_claims_paid <- util |>
  group_by(service_month) |>
  summarize(Total_Paid = sum(paid)) |>
  arrange(service_month)

claims_and_prems_per_month <- inner_join(total_claims_paid, prem, by = c("service_month" = "month"))

insurance_loss_ratio <- claims_and_prems_per_month |>
  mutate(Loss_Ratio_as_Percent = Total_Paid / premium * 100)

dlr_graph <- insurance_loss_ratio |>
  ggplot(aes(x = service_month, y = Loss_Ratio_as_Percent)) +
  geom_line(color = "navy") +
  labs(title = "Monthly Dental Loss Ratio", x = "Month", y = "Loss Ratio") +
  theme_minimal() +
  geom_hline(yintercept = 100, linetype = "dashed", color = "red") +
  scale_y_continuous(labels = scales::percent_format(scale = 1))

dlr_graph
```

```{r}
payments_by_lag <- util |>
  group_by(lag) |>
  summarise(total_paid = sum(paid), .groups = "drop") |>
  arrange(lag)

combined_df <- inner_join(payments_by_lag, cf_df, by = "lag") |>
  mutate(
    total_claims = total_paid / cf,
    IBNP = total_claims - total_paid
  )

# Map each service month's age to its completion factor (fully mature = 1)
cf_by_month <- c()
ages <- cum_tri$age
for (i in seq_along(ages)) {
  if (ages[i] >= 15) {
    cf_by_month <- c(cf_by_month, 1)
  } else {
    cf_by_month <- c(cf_by_month, cf_df$cf[ages[i] + 1])
  }
}

cf_loss_ratio <- cum_tri |>
  mutate(
    completion_factor = cf_by_month,
    paid_to_date = `15`,
    Ultimate = paid_to_date / completion_factor
  ) |>
  left_join(prem, by = c("service_month" = "month")) |>
  mutate(CF_lossratio = Ultimate / premium) |>
  select(service_month, CF_lossratio)

cf_loss_ratio_graph <- cf_loss_ratio |>
  ggplot(aes(x = service_month, y = CF_lossratio)) +
  geom_line(color = "purple") +
  labs(title = "Completion Factor Method: Adjusted Loss Ratio", x = "Month", y = "Loss Ratio") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )

cf_loss_ratio_graph
```

------------------------------------------------------------------------

## Method 2: Bornhuetter-Ferguson

```{r}
total_claims_paid <- util |>
  group_by(service_month) |>
  summarize(Total_Paid = sum(paid)) |>
  arrange(service_month) # earliest to latest service month

claims_and_prems_per_month <- inner_join(total_claims_paid, prem, by = c("service_month" = "month"))

insurance_loss_ratio <- claims_and_prems_per_month |>
  mutate(
    Loss_Ratio_as_Percent = Total_Paid / premium * 100,
    service_month = as.Date(service_month)
  )

# Expected proportion of ultimate claims paid by completion factor, per month
month_completion_factor <- c()
ages <- cum_tri$age
for (i in seq_along(ages)) {
  if (ages[i] >= 15) {
    month_completion_factor <- c(month_completion_factor, 1)
  } else {
    month_completion_factor <- c(month_completion_factor, cf_df$cf[ages[i] + 1])
  }
}

cum_tri <- cum_tri |>
  mutate(Exp_Proportion_Paid_by_CF = month_completion_factor)

# A priori expected loss ratio, estimated from fully mature (fully-developed) months only
fully_mature_months <- cum_tri |>
  filter(Exp_Proportion_Paid_by_CF == 1) |>
  select(service_month)

mature_loss_ratios <- insurance_loss_ratio |>
  filter(service_month %in% fully_mature_months$service_month)

a_priori_expected_loss_ratio <- sum(mature_loss_ratios$Loss_Ratio_as_Percent) /
  length(mature_loss_ratios$Loss_Ratio_as_Percent) * 0.01

# Step 1: % of ultimate claims still unreported at each month's current age
percent_unreported <- 1 - month_completion_factor
cum_tri <- cum_tri |>
  mutate(`% Unreported` = percent_unreported)

# Step 2: expected ultimate claims $ = a priori ELR x premium
cum_tri <- cum_tri |>
  left_join(prem, by = c("service_month" = "month")) |>
  mutate(expected_ultimate_claims_dollar_amt = a_priori_expected_loss_ratio * premium) |>
  select(-premium)

# Step 3: BF ultimate = paid-to-date + (expected ultimate x % unreported)
cum_tri <- cum_tri |>
  mutate(
    paid_to_date = mapply(function(sm, ac) {
      col <- as.character(min(ac, 15))
      cum_tri[cum_tri$service_month == sm, col, drop = TRUE]
    }, service_month, age),
    Ultimate = paid_to_date + (expected_ultimate_claims_dollar_amt * `% Unreported`)
  )

# Step 4: resulting loss ratio and chart
loss_ratio_by_BF <- cum_tri |>
  left_join(prem, by = c("service_month" = "month")) |>
  mutate(BF_lossratio = Ultimate / premium) |>
  select(service_month, BF_lossratio)

BF_lossratio_graph <- loss_ratio_by_BF |>
  ggplot(aes(x = service_month, y = BF_lossratio)) +
  geom_line(color = "navy") +
  labs(title = "Bornhuetter-Ferguson Method: Adjusted Loss Ratio", x = "Month", y = "Loss Ratio") +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )

BF_lossratio_graph
```

------------------------------------------------------------------------

## Method 3: Projected Paid Lag

```{r}
lag_cols <- as.character(0:15)

# Step 1: incremental % of ultimate claims paid at each lag,
# benchmarked against mature (15+ month old) service months
mature <- cum_tri |> filter(age >= 15)

lag_inc_pcts <- sapply(lag_cols, function(l) {
  prev_l <- as.character(as.integer(l) - 1)
  if (l == "0") {
    mean(mature[["0"]] / mature[["15"]], na.rm = TRUE)
  } else {
    mean((mature[[l]] - mature[[prev_l]]) / mature[["15"]], na.rm = TRUE)
  }
})

lag_pct_df <- data.frame(
  lag = as.integer(lag_cols),
  inc_pct = lag_inc_pcts
) |>
  mutate(cum_pct = cumsum(inc_pct))

print(lag_pct_df)

# Step 2: project ultimate claims per service month
# ultimate = paid-so-far / cumulative % expected paid by that age
plag_estimates <- cum_tri |>
  mutate(age_capped = pmin(age, 15)) |>
  left_join(lag_pct_df |> select(lag, cum_pct), by = c("age_capped" = "lag")) |>
  mutate(
    paid_so_far = mapply(function(sm, ac) {
      cum_tri[cum_tri$service_month == sm, as.character(ac), drop = TRUE]
    }, service_month, age_capped),
    projected_ultimate = ifelse(cum_pct > 0, paid_so_far / cum_pct, paid_so_far)
  ) |>
  select(service_month, paid_so_far, cum_pct, projected_ultimate)

plag_estimates <- plag_estimates |>
  left_join(prem, by = c("service_month" = "month")) |>
  mutate(IBNP = projected_ultimate - paid_so_far) |>
  select(service_month, paid_so_far, cum_pct, projected_ultimate, IBNP, premium)

# Step 3: resulting loss ratio
loss_ratio_plag <- plag_estimates |>
  mutate(lr_plag = projected_ultimate / premium) |>
  rename(month = service_month) |>
  arrange(month)

# Step 4: chart
ggplot(loss_ratio_plag, aes(x = month, y = lr_plag)) +
  geom_line(color = "#1a9641", linewidth = 1.1) +
  geom_point(color = "#1a9641", size = 1.5) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", alpha = 0.6) +
  labs(
    title = "Projected Paid Lag Method: Adjusted Loss Ratio",
    x = NULL,
    y = "Loss Ratio"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

------------------------------------------------------------------------

## Conclusion

Across all three methods, the mature-month loss ratios point to an a
priori expected loss ratio of roughly **72%**. Because premium was
growing quickly during the experience period, pure loss-development
approaches (like Completion Factor) are less reliable for the most
recent, least-mature months. The **Bornhuetter-Ferguson method** was
ultimately recommended, since it blends the a priori expected loss
ratio with actual paid claims and is more robust when recent periods
are still under-developed.
